\documentclass[12pt,letterpaper]{refart}

\usepackage{amsmath}                         % American Mathematical Society
\usepackage{bm}                              % bold math
\usepackage{braket}                          % Dirac notation
\usepackage{fancyvrb}                        % verbatim function
\usepackage[usenames,dvipsnames]{color}      % Periwinkle
\usepackage{listings}                        % typeset computer code
\usepackage{fancyhdr}                        % fancy page top and bottom
\usepackage{palatino}                        % cool font 
\usepackage{newpxtext}                       % associated math font for palatino
\usepackage{newpxmath}                       % associated math font for palatino
\usepackage[T1]{fontenc}                     % unicode catch-all

\usepackage[sort&compress,square,numbers]{natbib} % bibliography       
\usepackage{natmove}                              % bibliography 
\usepackage[colorlinks=true,
	citecolor=blue,
	linkcolor=blue,
	urlcolor=blue,
	pagebackref=false]{hyperref}             % hyperrefs
	
\renewcommand{\headrulewidth}{0.4pt} % line at top of page; 0.4pt typical
\renewcommand{\footrulewidth}{0.4pt} % line at bottom of page; 0.4pt typical
\newcommand{\myparagraph}[1]{\paragraph{#1 ---\hspace{-0.125in}}}

\newcommand{\VerbFcn}[1]{{\texttt{#1}}}
\newcommand{\VerbCmd}[1]{{\bf #1}}

\setlength{\parindent}{0.25in}
\setlength\papermarginwidth{1.50in}
\settextfraction{1.00}

\begin{document}
\sloppy

\title{
  The UniDyn \emph{Mathematica} package:
  Self-derived unitary operator rotations in quantum mechanics}
\author{
	John A. Marohn\thanks{jam99@cornell.edu} \\ 
	{\footnotesize Dept. of Chemistry and Chemical Biology} \\ 
	{\footnotesize Cornell University, Ithaca, NY 14851-1301, USA}
}
\maketitle

\begin{abstract}
\input{UniDyn-doc--abstract.tex}  
\end{abstract}

\pagestyle{fancy}  % page header and footers

	\lhead{\textsf{UniDyn \emph{Mathematica} package}}
	\rhead{\textsf{John A.\ Marohn}}
	\cfoot{\thepage}
	

\lstset{ %
% language=Mathematica,      % the language of the code -- buggy  
keywords={
	Clear,
	True,False,
	Exp, Sin, Cos, Tan,
	Not,
	Flatten,
	StringJoin, ToString, Length,
	Module, Do, For,
	Map, Table, Position, Sort, Sequence, List,
	Return, 
	Dimensions
	},
keywordstyle={\sffamily\bfseries\footnotesize}, % fonts used for keywords
basicstyle={\sffamily\footnotesize},          % fonts that are used for everything else
xleftmargin=2em,
xrightmargin=-2em,
tabsize=2,                 % sets default tabsize to 2 spaces
captionpos=t,              % sets the caption-position to bottom
caption=\lstname,          % show the filename 
comment=[is]{(**}{**)},    % completely IGNORE text between "(**" and "**)" in the code 
escapeinside={(*@}{@*)},   % code between "(*@" and "@*)" will be converted to LaTeX
%
escapebegin=\color{Periwinkle}\hspace{-2em}\begin{minipage}{\linewidth},
escapeend=\end{minipage}, 
breakindent=-2em,
postbreak=\space,
showstringspaces=false,
breakatwhitespace=true,
%
framexleftmargin=2em,
framexrightmargin=-2em,
frame=lines,
frame=tb,
%
rangeprefix=(*~\ ,  % we will include only code between the lines "(*~ START ~*)" 
rangesuffix=\ ~*),  % ... and "(*~ END ~*)"; the "START" and "END" parts are specified below
columns=flexible,   % this is CRUCIAL; now function names just look like times new roman font....
%
literate={x\$var}{{\emph{x}}}1 
{i\$max\$sym}{{\ensuremath{i_{\mathrm{max}}}}}1
{j\$max\$sym}{{\ensuremath{j_{\mathrm{max}}}}}1
{n\$sym}{{\emph{n}}}1
{i\$sym}{{\emph{i}}}1
{j\$sym}{{\emph{j}}}1
{a\$sym}{{\ensuremath{a_0}}}1
{b\$sym}{{\ensuremath{b_0}}}1
{c\$sym}{{\ensuremath{c_0}}}1
{d\$sym}{{\ensuremath{d_0}}}1
{t\$sym}{{\ensuremath{t}}}1
{rho\$sym}{{\ensuremath{\rho}}}1
{rho\$sym\$0}{{\ensuremath{\rho_{0}}}}1
{Ix\$sym}{{\ensuremath{I_x}}}1
{Iy\$sym}{{\ensuremath{I_y}}}1
{Iz\$sym}{{\ensuremath{I_z}}}1
{Sx\$sym}{{\ensuremath{S_x}}}1
{Sy\$sym}{{\ensuremath{S_y}}}1
{Sz\$sym}{{\ensuremath{S_z}}}1
{omega\$sym}{{\ensuremath{\omega}}}1
{p\$sym}{{\emph{p}}}1
{a\$new\$sym}{{\ensuremath{a_{\mathrm{new}}}}}1
{p\$new\$sym}{{\ensuremath{p_{\mathrm{new}}}}}1
{->}{{\ensuremath{\: \rightarrow \:}}}1
{//.}{{\ensuremath{\: //. \:}}}1  % define this before the {/.} replacement or //. will come out funny
{/.}{{\ensuremath{\: /. \:}}}1
{/@}{{\ensuremath{\: /@ \:}}}1
{Print["}{{Print[``}}8
{"Op}{{``Op}}3
{["}{{[``}}2
{,"}{{,``}}3
{A\$sym}{{\emph{A}}}1
{B\$sym}{{\emph{B}}}1
{C\$sym}{{\emph{C}}}1
{D\$sym}{{\emph{D}}}1
{sigma\$1\$sym}{{\ensuremath{\sigma_1}}}1
{aR\$sym}{{\ensuremath{a^{\dagger}}}}1
{aL\$sym}{{\ensuremath{a}}}1
{P\$sym}{{\ensuremath{P}}}1
{Q\$sym}{{\ensuremath{Q}}}1
{H\$sym}{{\ensuremath{{\cal H}}}}1
{Sqrt[2]}{{\ensuremath{\sqrt{2}}}}1
{bR\$sym}{{\ensuremath{b^{\dagger}}}}1
{bL\$sym}{{\ensuremath{b}}}1
{Nop\$a}{{\ensuremath{N_a}}}1
{Nop\$b}{{\ensuremath{N_b}}}1
{\\[Alpha]}{{\ensuremath{\alpha}}}1
{\\[Beta]}{{\ensuremath{\beta}}}1
{\\[Delta]}{{\ensuremath{\Delta}}}1
{\\[Omega]}{{\ensuremath{\omega}}}1
{\\[Rho]\$0}{{\ensuremath{\rho_{0}}}}1
{\\[CapitalDelta]}{{\ensuremath{\Delta}}}1
{alpha\$sym}{{\ensuremath{\alpha}}}1
{beta\$sym}{{\ensuremath{\beta}}}1
{constant1}{{\ensuremath{c_1}}}1
{constant2}{{\ensuremath{c_2}}}1
{constant3}{{\ensuremath{c_3}}}1
{constant4}{{\ensuremath{c_4}}}1
{omega\$eff}{{\ensuremath{\omega_{\text{eff}}}}}1
{rho\$known}{{\ensuremath{\rho_{\text{known}}}}}1
{rho\$calc}{{\ensuremath{\rho_{\text{calc}}}}}1
{delta$\p\$sym}{{\ensuremath{\delta p}}}1
{delta$\q\$sym}{{\ensuremath{\delta q}}}1
}

\section{Introduction}
% ====================

\input{UniDyn-doc--intro.tex}

\section{The algorithm}
% ===============

\input{UniDyn-doc--algorithm.tex}

\section{The code}
% ============

The code consists of \verb+.m+ files and \emph{Mathematica} notebooks.
Each modular \verb+.m+ file implements a desired function, like non-commutative multiplication or commutation.
Associated with each \verb+.m+ file is a second \verb+.m+ file containing \emph{units tests}.
The code, at the time of writing, includes over 120 unit tests covering all the functions defined below.
More detailed applications of the code are given in separate \emph{Mathematica} notebooks and will be discussed in the next section.

Code was written in the literate-programming style, with \LaTeX{}-style text, equations, and tables interspersed with code in the form of commented-out text.
In this section we use \LaTeX{}'s \verb+lstinputlisting+ package to import the code and typeset it nicely.
This package also takes care of importing and un-commenting the embedded comments so they can be processed in \LaTeX.
The package moreover implements a long list of character substitutions, allowing the code listings below to have variables appear with subscripts and superscripts.

\subsection{Operators and scalars}
% ------------------------------------------

Our first task is to define functions that enable \emph{Mathematica} to distinguish between (non-commutative) operators and (commutative) scalars. 
This is done in the \verb+OpQ.m+ package, whose listing appears below. 
A variable is either a commutative \emph{scalar} or non-commutative \emph{operator}.
Operators have a \emph{phylum} and an \emph{order}.
Operators of different phyla commute.
The operator order is used for sorting.

\lstinputlisting[linerange=START-END,includerangemarker=false]{unidyn/OpQ.m}

Unit tests for the functions in the \verb+OpQ.m+ package are organized into a separate package, \verb+OpQ-tests.m+, whose listing follows. 
Unit testing is built on \emph{Mathematica}'s \verb+VerificationTest+ function.
When the return value of the \verb+VerificationTest+ function is passed to the \verb+TestReport+ in a \emph{Mathematica} notebook, the notebook displays a nice graphical report card.

\lstinputlisting[linerange=START-END,includerangemarker=false]{unidyn/OpQ-tests.m}

\section{Non-commutative multiplication}
% ---------------------------------------------------

We next define our own non-commutative multiplication function, \verb+Mult+.
This is done in the \verb+Mult.m+ package.
We provide a sorting function which rearranges operators being multiplied so that, if possible, operators of a higher \emph{phylum} appear in front of operators of a lower \emph{phylum}.
During this rearrangement, non-commuting operators are not allowed to pass each other.

\lstinputlisting[linerange=START-END,includerangemarker=false]{unidyn/Mult.m}

Extensive unit testing shows that \verb+Mult+ distributes over addition, is associative, and pulls scalars out in front as desired.
Testing is also carried out on the sorting function.

\lstinputlisting[linerange=START-END,includerangemarker=false]{unidyn/Mult-tests.m}

\section{Commutator}
% -------------------------

Define a commutator function which factors out scalars and has rules for simplifying more complex commutators, like and $[A B, C]$ and $[A, B C]$, involving three non-commuting operators.

\lstinputlisting[linerange=START-END,includerangemarker=false]{unidyn/Comm.m}

Extensive testing shows that the commutator function factors out scalars correctly and correctly simplifies over a dozen commutation identities, including the Jacobi identity.

\lstinputlisting[linerange=START-END,includerangemarker=false]{unidyn/Comm-tests.m}

\section{Spins}
% ----------------

Declare angular momentum operators for one spin $I_x$, $I_y$, and $I_z$ and specify the relevant commutation relations.
To speed up computation, each operator's commutation relations are stored as an \emph{upvalue} of the operator instead of as a \emph{downvalue} of the commutator function.
Additional product-operator simplification rules are defined if the spin is declared to have an angular momentum of $\ell = 1/2$.

\lstinputlisting[linerange=START-END,includerangemarker=false]{unidyn/Spins.m}

Test the commutation relations for one spin and for a system of two spins, one of them $\ell = 1/2$ and the other with $\ell$ unspecified.

\lstinputlisting[linerange=START-END,includerangemarker=false]{unidyn/Spins-tests.m}

\section{Harmonic oscillator}
% -----------------------------------

Declare the annihilation (or lowering) operator $a$ and the creation (or raising) operator $a^{\dagger}$ for a harmonic oscillator, and specify the relevant commutation relations.

\lstinputlisting[linerange=START-END,includerangemarker=false]{unidyn/Osc.m}

Check commutation relations among the harmonic-oscillator operators: $a$, $a^{\dagger}$, the number operator $N = a^{\dagger} a$, the unitless position operator $Q = (a^{\dagger} + a)/\sqrt{2}$, and the unitless momentum operator $P = i (a^{\dagger} - a)/\sqrt{2}$.
Confirm that harmonic oscillator can exist ``on top of'' spin operators and are sorted correctly.

\lstinputlisting[linerange=START-END,includerangemarker=false]{unidyn/Osc-tests.m}

\section{Unitary Evolution}
% --------------------------------

Implement the \verb+Evolver+ algorithm described above.
The function can be passed an option to print out intermediate variables, which is helpful for debugging.

\lstinputlisting[linerange=START-END,includerangemarker=false]{unidyn/Evolve.m}

Unit testing includes verifying that \verb+Evolver+ distributes over sums and factors our scalars correctly.
Unit testing also verifies that \emph{Mathematica's} differential equation solver is behaving as expected.
We are not worried about bugs in the solver. 
We are instead concerned about changes in function syntax.
The solver function's syntax changed between \emph{Mathematica} version 8 and 10 and could change again. 
We added an equation-solver unit test to flag a future syntax change.

We test the following cases of unitary evolution in the rotating frame:

\begin{enumerate}
\item $e^{-i\omega t I_z } I_x e^{+i \omega t I_z}$ \\
  free evolution of transverse magnetization in a magnetic field
\item $e^{-i \omega t I_x} I_z e^{+i \omega t I_x}$ \\
  on-resonance nutation of longitudinal (i.e., thermal-equilibrium) magnetization in the rotating frame
\item $e^{-i \omega t I_z} I_{+} e^{+i \omega t I_z}$ \\
  free evolution of complex transverse magnetization in a magnetic field 
\item $e^{-i d t I_z S_z} I_x e^{+i d t I_z S_z}$ \\
  evolution of transverse magnetization under a scalar coupling
\item $e^{-i (\Delta \, I_z + \omega \, I_x) t} I_z e^{+i (\Delta \, I_z + \omega \, I_x) t}$ \\
  \label{item:nutation}
 off-resonance nutation of longitudinal magnetization
\item $e^{- \omega t \frac{1}{2} (a^{\dagger} \, a + a \, a^{\dagger})} a e^{+ \omega t \frac{1}{2} (a^{\dagger} \, a + a \, a^{\dagger})}$  \\
  evolution of the lowering (and raising) operator under the harmonic oscillator Hamiltonian
 \item $e^{- \omega t \frac{1}{2} (a^{\dagger} \, a + a \, a^{\dagger})} Q e^{+ \omega t \frac{1}{2} (a^{\dagger} \, a + a \, a^{\dagger})}$ \\
 evolution of the position (and momentum) operator under the harmonic oscillator Hamiltonian
 \item $e^{-i \delta q \, P }  Q e^{+i \delta q \, P }$ \\
  a position kick
\item $e^{-i \delta q \, Q }  P e^{+i \delta q \, Q }$ \\
  a momentum kick
\end{enumerate}
The off-resonance spin nutation, case~\ref{item:nutation},  is an especially demanding test because it generates terms involving $I_x$, $I_y$, and $I_z$ operators from solutions to a third-order differential equation.

\lstinputlisting[linerange=START-END,includerangemarker=false]{unidyn/Evolve-tests.m}

% Cited References
% ================

\clearpage
\renewcommand*{\bibfont}{\raggedright\normalfont\small}
\bibliographystyle{bib/grants-CV}
\bibliography{bib/UniDyn-doc}


\end{document}